---
title: "NC DATA"
author: "Susan Simmons"
date: "3/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


##NC format data

Many enviromental data sets are downloaded in the .nc format. When in this format, we need to extract the data in order to analyze it.  In R, this process is fairly straight forward when using the raster and ncdf4 packages. 

```{r packages}
library(raster)
library(ncdf4)
```

## Read in .NC data

We can use the brick command to read in the .nc formatted data set.  You will see three files brought into R using this command (be sure to change directory and file name to your data set).

```{r pressure, echo=FALSE}
anom.r=brick('Q:/My Drive/“AI” and Environmental Statistics/Working Group/New Soil Moisture SST Data/NOAA_NCDC_ERSST_v5anom_011948-122021.nc')

Soil.r=brick('Q:/My Drive/“AI” and Environmental Statistics/Working Group/New Soil Moisture SST Data/NOAA_NCEP_CPC_GMSM_w_US_011948-122021.nc')

Borneo.r=brick('Q:/My Drive/“AI” and Environmental Statistics/Working Group/New Soil Moisture SST Data/NOAA_NCEP_CPC_GMSM_w_Borneo_011948-122021.nc')
```

To extract the data, we need to specify the Longitude and Latitute coordinates and how frequently the observations are recorded.  For example, the lon.pon and lat.pon create a sequence of the Longitude and Latitute locations.  The data are recorded every two degrees.  You will see this repeated 3 times (Longitude and Latitute coordinates for the the three .nc files).

```{r extract data}

###These coordinates are for the Sea Surface Anomalies.
lon.pon=rep(seq(124,290,by=2),each=46)
lat.pon=rep(seq(-30,60,by=2),84)
extract.pts <- cbind(lon.pon,lat.pon)
###These coordinates are for the soil moisture data.
lon.pon.s=rep(seq(-125,-67.5,by=0.5),each=54)
lat.pon.s=rep(seq(24,50.5,by=0.5),116)
extract.pts.s <- cbind(lon.pon.s,lat.pon.s)
### These coordinates are for the Borneo data.
lon.pon.b=rep(seq(107.5,119,by=0.5),each=26)
lat.pon.b=rep(seq(-4,8.5,by=0.5),24)
extract.pts.b <- cbind(lon.pon.b,lat.pon.b)

### Now all three data sets can be extracted using their coordinates.
anom.r2=extract(anom.r,extract.pts,method="simple")
soil.r2=extract(Soil.r,extract.pts.s,method="simple")
Borneo.r2=extract(Borneo.r,extract.pts.b,method="simple")
```
