---
title: "XGBoost for AI"
author: "Susan Simmons"
date: "3/1/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import data sets

The following code is used on the clustered data set (so need to import the csv files that include cluster information).

```{r reading in data sets}
library(raster)
library(ncdf4)
library(ggplot2)
library(fpc)
library(dplyr)
library(caret)
library(randomForest)
library(doParallel)
library(xgboost)
library(plyr)
library(iml)
library(Ckmeans.1d.dp)
library(SHAPforxgboost)
SST.clus=read.csv('Q:\\My Drive\\“AI” and Environmental Statistics\\Working Group\\New Soil Moisture SST Data\\SST_withclus.csv')
soil.new=read.csv('Q:\\My Drive\\“AI” and Environmental Statistics\\Working Group\\New Soil Moisture SST Data\\Soil.csv')
Borneo.new=read.csv('Q:\\My Drive\\“AI” and Environmental Statistics\\Working Group\\New Soil Moisture SST Data\\Borneo.csv')
```

## Feature creation

With algorithms such as XG-Boost, it is important to create features that provide information on the response variable.  The code below calculates means, variances, min and max information for each cluster in each time frame.

```{r feature creation}
SST.clus2=data.frame(SST.clus)
colnames(SST.clus2)[1]="Clust"
SST_var=SST.clus2[,-c(2:3)] %>% group_by(Clust)   %>% summarize_all(var)
SST_var=SST_var[,-1]
SST_mean=SST.clus2[,-c(2:3)] %>% group_by(Clust)   %>% summarize_all(mean)
SST_mean=SST_mean[,-1]
SST_max=SST.clus2[,-c(2:3)] %>% group_by(Clust)   %>% summarize_all(max)
SST_max=SST_max[,-1]
SST_min=SST.clus2[,-c(2:3)] %>% group_by(Clust)   %>% summarize_all(min)
SST_min=SST_min[,-1]

```

## X matrix for sea surface data
The following x matrix will be repeated for each location (using a training data set of 1948-2013 or 792 months).

The rol.lag is number of months to "roll-up".  The for.ahead indicates how far out to forecast (for example, if for.ahead is 1, then forecasting next month.)
```{r data manipulation}
##rol.lag is how many lags to "roll up to consider in analysis"
rol.lag=3
#for.ahead is the number of months ahead to forecast (if for.ahead=1, then forecasting next month)
for.ahead=3

###Using the clustered data
### Need to use years 1948-2013 as training
### corresponds to columns 1-792(test is 793-888)

## Create X matrix
## mean of cluster for number of lags requested (mean)
## variance of clusters for number of lags requested (median)
## Change from previous number of lags months (max[mean] - min[mean])
## Change in deviations from previous numbers of lags (current median std/old median std)
## Month being forecasted
## Soil info lagged
x.train=matrix(nrow=(792-2*rol.lag-for.ahead+1),ncol=300)

for (i in (rol.lag):(792-rol.lag-for.ahead)){
        temp1=SST_mean[,(i+1):(i+rol.lag)]
        temp1.1=apply(temp1,1,mean)
        temp2=SST_var[,(i+1):(i+rol.lag)]
        temp2.2=apply(temp2,1,median)
        temp3.1=SST_max[,(i+1):(i+rol.lag)]
        temp3.max=apply(temp3.1,1,max)
        temp3.2=SST_min[,(i-rol.lag+1):(i)]
        temp3.min=apply(temp3.2,1,min)
        temp3.3=temp3.max-temp3.min
        temp4.2=SST_var[,(i-rol.lag+1):(i)]
        temp4.old=apply(temp4.2,1,median)
        temp4.4=temp2.2/temp4.old
        x.train[i-rol.lag+1,]=c(temp1.1,temp2.2,temp3.3,temp4.4)
}
## For month information
months.mat=kronecker(rep(1,100),diag(12))
months.mat1=months.mat[1:nrow(x.train),]
x.train=cbind(x.train,months.mat1)
########################################
#######################################
#x.train is repeated for every soil location...



```

Now for the test set...

```{r}
x.test=matrix(nrow=92,ncol=300)
for (i in (792+rol.lag-4):(888-rol.lag-for.ahead)){
        temp1=SST_mean[,(i+1):(i+rol.lag)]
        temp1.1=apply(temp1,1,mean)
        temp2=SST_var[,(i+1):(i+rol.lag)]
        temp2.2=apply(temp2,1,median)
        temp3.1=SST_max[,(i+1):(i+rol.lag)]
        temp3.max=apply(temp3.1,1,max)
        temp3.2=SST_min[,(i-rol.lag+1):(i)]
        temp3.min=apply(temp3.2,1,min)
        temp3.3=temp3.max-temp3.min
        temp4.2=SST_var[,(i-rol.lag+1):(i)]
        temp4.old=apply(temp4.2,1,median)
        temp4.4=temp2.2/temp4.old
        x.test[i-792+2,]=c(temp1.1,temp2.2,temp3.3,temp4.4)
}
## For month information
months.mat=kronecker(rep(1,9),diag(12))
months.mat1=months.mat[1:nrow(x.test),]
x.test=cbind(x.test,months.mat1)

```


Since we are focusing on corn belt region, need to subset the soil moisture data.


```{r Corn belt}

corn.soil=soil.new %>% filter(between(Lon,-102,-80) & between(Lat,35.5,49))
corn.soil=corn.soil[,-c(1:2)]

corn.soil2=apply(corn.soil,2,mean)

```



## Run the XGBoost model

The following code runs the XGBoost model (this will take a while). The Shapley matrix is storing ONLY the Shapley values for May (this is where we are focusing our attention).


```{r,results='hide'}

all.pred.y=matrix(nrow=nrow(corn.soil),ncol=92)
shap.May=matrix(nrow=(8*nrow(corn.soil)),ncol=315)
y.test.all=matrix(nrow=nrow(corn.soil),ncol=92)
tt=12*seq(0,7)+1

## Need to do for k=1 first
k=1
y.train=vector(length=nrow(x.train))
soil.lag=matrix(nrow=784,ncol=3)
for (i in (rol.lag):(792-rol.lag-for.ahead))
    {soil.lag[(i-rol.lag+1),]=as.numeric(unlist(corn.soil[k,(i+1):(i+rol.lag)]))
    y.train[(i-rol.lag+1)]=corn.soil[k,i+rol.lag+for.ahead]
    }
 x.train.all=cbind(x.train,soil.lag)
 
soil.lag=matrix(nrow=92,ncol=rol.lag)
y.test=vector(length=nrow(x.test))

for (i in (792+rol.lag-4):(888-rol.lag-for.ahead))
    {soil.lag[i-rol.lag-792+1,]=as.numeric(unlist(corn.soil[k,(i+1):(i+rol.lag)]))
    y.test[(i-rol.lag-792+1+4)]=corn.soil[k,i+rol.lag+for.ahead]
    y.test.all[k,i-792-rol.lag+1+4]=corn.soil[k,i+rol.lag+for.ahead]
    }
 x.test.all=cbind(x.test,soil.lag)


dtrain <- xgb.DMatrix(data.matrix(x.train.all),label = y.train)
dvalid <- xgb.DMatrix(data.matrix(x.test.all),label = y.test)
params <- list(
  objective = "reg:squarederror",
  learning_rate = 0.05,
  subsample = 0.9,
  colsample_bynode = 1,
  reg_lambda = 1,
  max_depth = 5
)
set.seed(58914)
fit_xgb <- xgb.train(
  params,
  data = dtrain,
  watchlist = list(valid = dvalid),
  early_stopping_rounds = 20,
  print_every_n = 100,
  nrounds = 1000 # early stopping
) 
pred.xgb=predict(fit_xgb,x.test.all)
pred.xgb2=predict(object = fit_xgb, newdata = x.test.all, predcontrib = TRUE)

all.pred.y[k,]=pred.xgb
shap.May=pred.xgb2[tt,]

## Now for all other locations in corn belt

for (k in 2:nrow(corn.soil)){   
y.train=vector(length=nrow(x.train))
soil.lag=matrix(nrow=784,ncol=3)
for (i in (rol.lag):(792-rol.lag-for.ahead))
    {soil.lag[(i-rol.lag+1),]=as.numeric(unlist(corn.soil[k,(i+1):(i+rol.lag)]))
    y.train[(i-rol.lag+1)]=corn.soil[k,i+rol.lag+for.ahead]
    }
 x.train.all=cbind(x.train,soil.lag)
 
soil.lag=matrix(nrow=92,ncol=rol.lag)
y.test=vector(length=nrow(x.test))

for (i in (792+rol.lag-4):(888-rol.lag-for.ahead))
    {soil.lag[i-rol.lag-792+1,]=as.numeric(unlist(corn.soil[k,(i+1):(i+rol.lag)]))
    y.test[(i-rol.lag-792+1+4)]=corn.soil[k,i+rol.lag+for.ahead]
    y.test.all[k,i-792-rol.lag+1+4]=corn.soil[k,i+rol.lag+for.ahead]
    }
 x.test.all=cbind(x.test,soil.lag)


dtrain <- xgb.DMatrix(data.matrix(x.train.all),label = y.train)
dvalid <- xgb.DMatrix(data.matrix(x.test.all),label = y.test)
params <- list(
  objective = "reg:squarederror",
  learning_rate = 0.05,
  subsample = 0.9,
  colsample_bynode = 1,
  reg_lambda = 1,
  max_depth = 5
)
set.seed(58914)
fit_xgb <- xgb.train(
  params,
  data = dtrain,
  watchlist = list(valid = dvalid),
  early_stopping_rounds = 20,
  print_every_n = 100,
  nrounds = 10000 # early stopping
) 
pred.xgb=predict(fit_xgb,x.test.all)
pred.xgb2=predict(object = fit_xgb, newdata = x.test.all, predcontrib = TRUE)

all.pred.y[k,]=pred.xgb
temp=pred.xgb2[tt,]
shap.May=rbind(shap.May,temp)
}

```

## Plots

Now for the plots

```{r plots}

temp=seq(0,1223)

year.index=seq(1,8)
for (i in year.index)
 {temp2=8*temp+i  ### +1 is 2014, +2 is 2015..
  shap2=shap.May[temp2,]
  shap3=apply(shap2,2,mean)
  if (i==1) {dat=cbind(seq(1,75),as.numeric(shap3[1:75]))}
  if (i>1) {dat=cbind(dat,as.numeric(shap3[1:75]))}
  }
colnames(dat)=c("Clust","Shap2014","Shap2015","Shap2016","Shap2017","Shap2018","Shap2019","Shap2020","Shap2021")
dat=data.frame(dat)
SST.clus2=data.frame(SST.clus)
colnames(SST.clus2)[1]="Clust"
SSTvis=left_join(SST.clus2[,1:3],dat,by="Clust")
SSTvis=data.frame(SSTvis)

ggplot(SSTvis,aes(x=Lon,y=Lat)) +               geom_point(data=SSTvis,aes(x=Lon,y=Lat,col=Shap2014),size=1.2) +
 scale_color_gradient2(low = "blue", mid = "white", high = "red", space = "Lab" ) +
geom_hline(yintercept=c(0)) +
  theme_classic()+labs(title="May 2014")

ggplot(SSTvis,aes(x=Lon,y=Lat)) +               geom_point(data=SSTvis,aes(x=Lon,y=Lat,col=Shap2015),size=1.2) +
 scale_color_gradient2(low = "blue", mid = "white", high = "red", space = "Lab" ) +
geom_hline(yintercept=c(0)) +
  theme_classic()+labs(title="May 2015")

ggplot(SSTvis,aes(x=Lon,y=Lat)) +               geom_point(data=SSTvis,aes(x=Lon,y=Lat,col=Shap2016),size=1.2) +
 scale_color_gradient2(low = "blue", mid = "white", high = "red", space = "Lab" ) +
geom_hline(yintercept=c(0)) +
  theme_classic()+labs(title="May 2016")

ggplot(SSTvis,aes(x=Lon,y=Lat)) +               geom_point(data=SSTvis,aes(x=Lon,y=Lat,col=Shap2017),size=1.2) +
 scale_color_gradient2(low = "blue", mid = "white", high = "red", space = "Lab" ) +
geom_hline(yintercept=c(0)) +
  theme_classic()+labs(title="May 2017")

ggplot(SSTvis,aes(x=Lon,y=Lat)) +               geom_point(data=SSTvis,aes(x=Lon,y=Lat,col=Shap2018),size=1.2) +
 scale_color_gradient2(low = "blue", mid = "white", high = "red", space = "Lab" ) +
geom_hline(yintercept=c(0)) +
  theme_classic()+labs(title="May 2018")

ggplot(SSTvis,aes(x=Lon,y=Lat)) +               geom_point(data=SSTvis,aes(x=Lon,y=Lat,col=Shap2019),size=1.2) +
 scale_color_gradient2(low = "blue", mid = "white", high = "red", space = "Lab" ) +
geom_hline(yintercept=c(0)) +
  theme_classic()+labs(title="May 2019")

ggplot(SSTvis,aes(x=Lon,y=Lat)) +               geom_point(data=SSTvis,aes(x=Lon,y=Lat,col=Shap2020),size=1.2) +
 scale_color_gradient2(low = "blue", mid = "white", high = "red", space = "Lab" ) +
geom_hline(yintercept=c(0)) +
  theme_classic()+labs(title="May 2020")

ggplot(SSTvis,aes(x=Lon,y=Lat)) +               geom_point(data=SSTvis,aes(x=Lon,y=Lat,col=Shap2021),size=1.2) +
 scale_color_gradient2(low = "blue", mid = "white", high = "red", space = "Lab" ) +
geom_hline(yintercept=c(0)) +
  theme_classic()+labs(title="May 2021")


year.plot=c(rep(2014,1224*8), rep(2015,1224*12),rep(2016,1224*12),rep(2017,1224*12),rep(2018,1224*12),rep(2019,1224*12),rep(2020,1224*12),rep(2021,1224*12))
y.all=data.frame(cbind(c(all.pred.y),c(y.test.all),year.plot))

colnames(y.all)=c("Predicted","Actual","Year")
ggplot(y.all,aes(x=Actual,y=Predicted)) +           geom_point(data=y.all,aes(x=Actual,y=Predicted,color=Year),size=0.5,shape=20) + scale_color_gradient(low = "yellow", high = "darkgreen", space = "Lab" )+
  theme_classic()+labs(title="All Months")


year.plot=c(rep(2014,1224), rep(2015,1224),rep(2016,1224),rep(2017,1224),rep(2018,1224),rep(2019,1224),rep(2020,1224),rep(2021,1224))

only.may=12*seq(0,7)+1
all.1=all.pred.y[,only.may]
y.test.1=y.test.all[,only.may]
y.all=data.frame(cbind(c(all.1),c(y.test.1),year.plot))

colnames(y.all)=c("Predicted","Actual","Year")
ggplot(y.all,aes(x=Actual,y=Predicted)) +           geom_point(data=y.all,aes(x=Actual,y=Predicted,color=Year),size=0.5,shape=20) + scale_color_gradient(low = "yellow", high = "darkgreen", space = "Lab" )+
  theme_classic()+labs(title="May")

mspe=vector(length=8)
for (i in year.index){
  if (i==1) {one.year=seq(1,8)}
  if (i>1)  {one.year=8+(i-2)*12 + seq(1,12)}
  all.1=all.pred.y[,one.year]
  y.test.1=y.test.all[,one.year]
  temp=mean((c(y.test.1)-c(all.1))^2)
  mspe[i]=temp}

mean((c(y.test.all)-c(all.pred.y))^2)
1-sum((y.test.all-all.pred.y)^2)/sum((y.test.all-mean(y.test.all))^2)


##May
tt=12*seq(0,7)+5
y.test.may=y.test.all[,tt]
all.pred.may=all.pred.y[,tt]
mean((c(y.test.may)-c(all.pred.may))^2)
1-sum((y.test.may-all.pred.may)^2)/sum((y.test.may-mean(y.test.may))^2)


```
```{r different areas}
##Trying a few single areas..
#k=1058 NE area
k=1058
i=1 ##May 2014
dat=cbind(seq(1,75),as.numeric(shap.May[(i+(k-1)*8),1:75]))

dat=data.frame(dat)
colnames(dat)=c("Clust","Shap")
SST.clus2=data.frame(SST.clus)
colnames(SST.clus2)[1]="Clust"
SSTvis=left_join(SST.clus2[,1:3],dat,by="Clust")
SSTvis=data.frame(SSTvis)

ggplot(SSTvis,aes(x=Lon,y=Lat)) +               geom_point(data=SSTvis,aes(x=Lon,y=Lat,col=Shap),size=1.2) +
 scale_color_gradient2(low = "blue", mid = "white", high = "red", space = "Lab" ) +
geom_hline(yintercept=c(0)) +
  theme_classic()+labs(title="May 2014")

```

